---
title: "Getting Started with siconfir"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with siconfir}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## What is siconfir?

**siconfir** is an R package that provides direct access to fiscal data from
[SICONFI](https://siconfi.tesouro.gov.br/) (Accounting and Fiscal Information
System for the Brazilian Public Sector) through the Brazilian National
Treasury's API.

It allows you to query data from all Brazilian states and municipalities,
including fiscal reports such as RREO, RGF, DCA, and the Accounting Balances
Matrix (MSC).

## Installation

```{r install}
# Install from CRAN (when available)
install.packages("siconfir")

# Or the development version from GitHub
# install.packages("pak")
pak::pak("StrategicProjects/siconfir")
```

## Basic queries

### Function names and English aliases

Every function is available in two forms: an English interface with English
parameter names, and a Portuguese interface matching the original API parameter
names. Both produce identical results -- use whichever you prefer:

| English function | Portuguese function |
|---|---|
| `get_entities()` | `get_entes()` |
| `get_annexes()` | `get_anexos()` |
| `get_annual_accounts()` | `get_dca()` |
| `get_delivery_status()` | `get_extrato()` |
| `get_budget_report()` | `get_rreo()` |
| `get_fiscal_report()` | `get_rgf()` |
| `get_msc_equity()` | `get_msc_patrimonial()` |
| `get_msc_budget()` | `get_msc_orcamentaria()` |
| `get_msc_control()` | `get_msc_controle()` |

The English functions also translate the **parameter names**. For example,
`fiscal_year` maps to `an_exercicio`, `entity_id` to `id_ente`, and so on.
See each function's help page for the full mapping.

Note that while function and parameter *names* are in English, some parameter
*values* must still be in Portuguese because they come from the API itself
(e.g., `appendix = "RREO-Anexo 01"`).

### List government entities

The most common starting point is to list the available entities. You will need
the IBGE code (`cod_ibge`) for all other queries.

```{r entes}
library(siconfir)
library(dplyr)

entes <- get_entities()
entes
```

You can filter by state, region, or government sphere:

```{r entes-filter}
# State capitals
capitals <- entes |>
  filter(capital == 1)

# Municipalities in the state of Pernambuco
municipalities_pe <- entes |>
  filter(uf == "PE", esfera == "M")
```

### Query the RREO (Budget Execution Summary Report)

The RREO is published bimonthly. Use `get_budget_report()` (or `get_rreo()`):

```{r rreo}
rreo <- get_budget_report(
  fiscal_year = 2023,
  period      = 6,
  report_type = "RREO",
  appendix    = "RREO-Anexo 01",
  sphere      = "E",
  entity_id   = 26 # Pernambuco
)
```

### Query the RGF (Fiscal Management Report)

The RGF is published every four months:

```{r rgf}
rgf <- get_fiscal_report(
  fiscal_year = 2023,
  periodicity = "Q",
  period      = 3,
  report_type = "RGF",
  appendix    = "RGF-Anexo 01",
  sphere      = "E",
  branch      = "E",
  entity_id   = 26
)
```

### Query the DCA (Annual Accounts Declaration)

The DCA contains annual balance sheet data:

```{r dca}
dca <- get_annual_accounts(fiscal_year = 2023, entity_id = 26)
```

### Check delivery status

To find out which reports an entity has submitted:

```{r extrato}
status <- get_delivery_status(entity_id = 26, year = 2023)
```

### Query the MSC (Accounting Balances Matrix)

The MSC provides detailed accounting data, split into three segments:
equity/assets (classes 1--4), budgetary (5--6), and control (7--8).

```{r msc}
# Equity/asset accounts
msc_equity <- get_msc_equity(
  entity_id     = 26,
  year          = 2023,
  month         = 12,
  matrix_type   = "MSCC",
  account_class = 1,
  value_type    = "ending_balance"
)

# Budgetary accounts
msc_budget <- get_msc_budget(
  entity_id     = 26,
  year          = 2023,
  month         = 12,
  matrix_type   = "MSCC",
  account_class = 6,
  value_type    = "period_change"
)
```

## Caching

The package maintains an in-memory cache for the duration of the R session.
If you call the same function with the same parameters, data will be returned
from the cache without hitting the API again.
  
To clear the cache:

```{r cache}
siconfir_clear_cache()
```

You can also disable the cache for a specific call:

```{r no-cache}
entes <- get_entes(use_cache = FALSE)
```

## Pagination

The SICONFI API returns a maximum of 5,000 rows per page. **siconfir** handles
pagination automatically, following the `hasMore` / `offset` pattern until all
data has been retrieved. Informative messages indicate how many pages were
fetched and how many rows were returned.

## Report appendices reference table

Use `get_anexos()` to query the reference table listing all available appendices
by report type and government sphere:

```{r anexos}
anexos <- get_anexos()

# Appendices available for the municipal RREO
anexos |>
  filter(esfera == "M", str_detect(anexo, "RREO"))
```

## Further information

- [SICONFI API documentation](https://apidatalake.tesouro.gov.br/docs/siconfi/)
- [SICONFI website](https://siconfi.tesouro.gov.br/)
