---
title: "Parameter Mapping Reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parameter Mapping Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

The **siconfir** package wraps the SICONFI API from Brazil's National Treasury.
Because the API is natively in Portuguese, the package offers two parallel
interfaces:

- **English interface** -- English function names *and* English parameter names
  (e.g., `get_budget_report(fiscal_year = 2023)`)
- **Portuguese interface** -- names matching the original API query parameters
  (e.g., `get_rreo(an_exercicio = 2023)`)

Both interfaces call the same underlying API endpoints and return identical
tibbles. This vignette provides a complete reference of all parameter mappings,
valid values, and side-by-side examples for every function.

> **Note:** While function and parameter *names* are translated, the parameter
> *values* and the *response column names* remain in Portuguese because they
> come directly from the API (e.g., `appendix = "RREO-Anexo 01"`,
> response column `cod_ibge`).

## Master parameter mapping table

The table below lists every API query parameter, its English alias, and which
functions use it.

| API parameter | English name | Type | Used in |
|---|---|---|---|
| `an_exercicio` | `fiscal_year` | Integer | DCA, RREO, RGF |
| `id_ente` | `entity_id` | Integer | DCA, extrato, RREO, RGF, MSC* |
| `no_anexo` | `appendix` | Character | DCA, RREO, RGF |
| `an_referencia` | `year` | Integer | extrato, MSC* |
| `me_referencia` | `month` | Integer | MSC* |
| `nr_periodo` | `period` | Integer | RREO, RGF |
| `co_tipo_demonstrativo` | `report_type` | Character | RREO, RGF |
| `co_esfera` | `sphere` | Character | RREO, RGF |
| `co_poder` | `branch` | Character | RGF |
| `in_periodicidade` | `periodicity` | Character | RGF |
| `co_tipo_matriz` | `matrix_type` | Character | MSC* |
| `classe_conta` | `account_class` | Integer | MSC* |
| `id_tv` | `value_type` | Character | MSC* |

\* MSC = `get_msc_equity()` / `get_msc_budget()` / `get_msc_control()`

---

## Function-by-function reference

### 1. Entities -- `get_entities()` / `get_entes()`

**Endpoint:** `/entes`

No required parameters. Returns all registered government entities.

```{r}
library(siconfir)

# English
entes <- get_entities()

# Portuguese (equivalent)
entes <- get_entes()
```

**Key response columns:** `cod_ibge`, `ente`, `capital`, `regiao`, `uf`,
`esfera`, `populacao`, `co_cnpj`.

---

### 2. Annexes -- `get_annexes()` / `get_anexos()`

**Endpoint:** `/anexos-relatorios`

No required parameters. Returns the reference table of available report
appendices.

```{r}
# English
anexos <- get_annexes()

# Portuguese (equivalent)
anexos <- get_anexos()
```

**Key response columns:** `esfera`, `demonstrativo`, `anexo`.

Use this to discover valid `appendix` / `no_anexo` values for other functions:

```{r}
library(dplyr)

# What appendices exist for the municipal RREO?
get_annexes() |>
  filter(esfera == "M", demonstrativo == "RREO")
```

---

### 3. Annual accounts -- `get_annual_accounts()` / `get_dca()`

**Endpoint:** `/dca`

| English param | Portuguese param | Required | Description |
|---|---|---|---|
| `fiscal_year` | `an_exercicio` | Yes | Fiscal year (e.g., `2023`) |
| `entity_id` | `id_ente` | Yes | IBGE code of the entity |
| `appendix` | `no_anexo` | No | Filter by appendix name |

```{r}
# English
dca <- get_annual_accounts(
  fiscal_year = 2023,
  entity_id   = 26  # Pernambuco
)

# Portuguese (equivalent)
dca <- get_dca(
  an_exercicio = 2023,
  id_ente      = 26
)

# With optional appendix filter
dca_ab <- get_annual_accounts(
  fiscal_year = 2023,
  entity_id   = 26,
  appendix    = "DCA-Anexo I-AB"
)
```

**Key response columns:** `exercicio`, `instituicao`, `cod_ibge`, `uf`,
`anexo`, `rotulo`, `coluna`, `cod_conta`, `conta`, `valor`, `populacao`.

---

### 4. Delivery status -- `get_delivery_status()` / `get_extrato()`

**Endpoint:** `/extrato_entregas`

| English param | Portuguese param | Required | Description |
|---|---|---|---|
| `entity_id` | `id_ente` | Yes | IBGE code of the entity |
| `year` | `an_referencia` | Yes | Reference year |

```{r}
# English
status <- get_delivery_status(
  entity_id = 26,
  year      = 2023
)

# Portuguese (equivalent)
status <- get_extrato(
  id_ente       = 26,
  an_referencia = 2023
)
```

**Key response columns:** `exercicio`, `cod_ibge`, `instituicao`,
`entregavel`, `periodo`, `periodicidade`, `status_relatorio`, `data_status`,
`forma_envio`, `tipo_relatorio`.

Useful for checking which reports have been submitted before querying them:

```{r}
library(dplyr)

# Which RREO periods did Pernambuco deliver in 2023?
get_delivery_status(entity_id = 26, year = 2023) |>
  filter(entregavel == "RREO") |>
  select(periodo, status_relatorio, data_status)
```

---

### 5. Budget Execution Summary Report -- `get_budget_report()` / `get_rreo()`

**Endpoint:** `/rreo`

| English param | Portuguese param | Required | Description |
|---|---|---|---|
| `fiscal_year` | `an_exercicio` | Yes | Fiscal year |
| `period` | `nr_periodo` | Yes | Bimester (1--6) |
| `report_type` | `co_tipo_demonstrativo` | Yes | `"RREO"` or `"RREO Simplificado"` |
| `appendix` | `no_anexo` | Yes | Appendix name |
| `sphere` | `co_esfera` | Yes | `"M"`, `"E"`, or `"U"` |
| `entity_id` | `id_ente` | Yes | IBGE code |

**Valid values for `report_type`:**

- `"RREO"` -- standard report (all entities)
- `"RREO Simplificado"` -- simplified report (municipalities under 50k)

**Valid values for `sphere`:**

- `"M"` -- municipalities
- `"E"` -- states (including Federal District)
- `"U"` -- federal government (union)

```{r}
# English -- 6th bimester for Pernambuco
rreo <- get_budget_report(
  fiscal_year = 2023,
  period      = 6,
  report_type = "RREO",
  appendix    = "RREO-Anexo 01",
  sphere      = "E",
  entity_id   = 26
)

# Portuguese (equivalent)
rreo <- get_rreo(
  an_exercicio          = 2023,
  nr_periodo            = 6,
  co_tipo_demonstrativo = "RREO",
  no_anexo              = "RREO-Anexo 01",
  co_esfera             = "E",
  id_ente               = 26
)

# Simplified RREO for a small municipality (Recife = 2611606)
rreo_simpl <- get_budget_report(
  fiscal_year = 2023,
  period      = 6,
  report_type = "RREO Simplificado",
  appendix    = "RREO-Anexo 01",
  sphere      = "M",
  entity_id   = 2611606
)
```

**Key response columns:** `exercicio`, `demonstrativo`, `periodo`,
`instituicao`, `cod_ibge`, `uf`, `populacao`, `anexo`, `coluna`,
`cod_conta`, `conta`, `valor`.

---

### 6. Fiscal Management Report -- `get_fiscal_report()` / `get_rgf()`

**Endpoint:** `/rgf`

| English param | Portuguese param | Required | Description |
|---|---|---|---|
| `fiscal_year` | `an_exercicio` | Yes | Fiscal year |
| `periodicity` | `in_periodicidade` | Yes | `"Q"` (four-monthly) or `"S"` (semi-annual) |
| `period` | `nr_periodo` | Yes | Period (1--3 for Q, 1--2 for S) |
| `report_type` | `co_tipo_demonstrativo` | Yes | `"RGF"` or `"RGF Simplificado"` |
| `appendix` | `no_anexo` | Yes | Appendix name |
| `sphere` | `co_esfera` | Yes | `"M"`, `"E"`, or `"U"` |
| `branch` | `co_poder` | Yes | Government branch code |
| `entity_id` | `id_ente` | Yes | IBGE code |

**Valid values for `branch` / `co_poder`:**

- `"E"` -- Executive
- `"L"` -- Legislative
- `"J"` -- Judiciary
- `"M"` -- Public Ministry
- `"D"` -- Public Defender

**Valid values for `periodicity` / `in_periodicidade`:**

- `"Q"` -- four-monthly (standard, 3 periods per year)
- `"S"` -- semi-annual (simplified RGF only, 2 periods per year)

```{r}
# English -- 3rd four-month period, executive branch, state level
rgf <- get_fiscal_report(
  fiscal_year = 2023,
  periodicity = "Q",
  period      = 3,
  report_type = "RGF",
  appendix    = "RGF-Anexo 01",
  sphere      = "E",
  branch      = "E",
  entity_id   = 26
)

# Portuguese (equivalent)
rgf <- get_rgf(
  an_exercicio          = 2023,
  in_periodicidade      = "Q",
  nr_periodo            = 3,
  co_tipo_demonstrativo = "RGF",
  no_anexo              = "RGF-Anexo 01",
  co_esfera             = "E",
  co_poder              = "E",
  id_ente               = 26
)

# Legislative branch
rgf_leg <- get_fiscal_report(
  fiscal_year = 2023,
  periodicity = "Q",
  period      = 3,
  report_type = "RGF",
  appendix    = "RGF-Anexo 01",
  sphere      = "E",
  branch      = "L",
  entity_id   = 26
)
```

**Key response columns:** `exercicio`, `periodo`, `periodicidade`,
`instituicao`, `cod_ibge`, `uf`, `co_poder`, `populacao`, `anexo`,
`coluna`, `cod_conta`, `conta`, `valor`.

---

### 7. MSC equity/asset accounts -- `get_msc_equity()` / `get_msc_patrimonial()`

**Endpoint:** `/msc_patrimonial`

| English param | Portuguese param | Required | Description |
|---|---|---|---|
| `entity_id` | `id_ente` | Yes | IBGE code |
| `year` | `an_referencia` | Yes | Reference year |
| `month` | `me_referencia` | Yes | Reference month (1--12) |
| `matrix_type` | `co_tipo_matriz` | Yes | `"MSCC"` or `"MSCE"` |
| `account_class` | `classe_conta` | Yes | `1`, `2`, `3`, or `4` |
| `value_type` | `id_tv` | Yes | Balance type (see below) |

**Valid values for `matrix_type` / `co_tipo_matriz`:**

- `"MSCC"` -- monthly aggregate matrix
- `"MSCE"` -- annual closing matrix

**Valid values for `value_type` / `id_tv`:**

- `"beginning_balance"` -- opening balance for the period
- `"ending_balance"` -- closing balance for the period
- `"period_change"` -- net movement during the period

**Valid values for `account_class` / `classe_conta`:**

- `1` -- Assets (Ativo)
- `2` -- Liabilities (Passivo)
- `3` -- Net equity variations -- additions (VPA)
- `4` -- Net equity variations -- deductions (VPD)

```{r}
# English -- assets at year-end
msc_assets <- get_msc_equity(
  entity_id     = 26,
  year          = 2023,
  month         = 12,
  matrix_type   = "MSCC",
  account_class = 1,
  value_type    = "ending_balance"
)

# Portuguese (equivalent)
msc_assets <- get_msc_patrimonial(
  id_ente        = 26,
  an_referencia  = 2023,
  me_referencia  = 12,
  co_tipo_matriz = "MSCC",
  classe_conta   = 1,
  id_tv          = "ending_balance"
)

# Liabilities movement during June
msc_liab <- get_msc_equity(
  entity_id     = 26,
  year          = 2023,
  month         = 6,
  matrix_type   = "MSCC",
  account_class = 2,
  value_type    = "period_change"
)
```

**Key response columns:** `tipo_matriz`, `cod_ibge`, `classe_conta`,
`conta_contabil`, `poder_orgao`, `fonte_recursos`, `exercicio`,
`mes_referencia`, `valor`, `natureza_conta`, `tipo_valor`.

---

### 8. MSC budgetary accounts -- `get_msc_budget()` / `get_msc_orcamentaria()`

**Endpoint:** `/msc_orcamentaria`

Same parameters as MSC equity, except:

| English param | Portuguese param | Required | Valid values |
|---|---|---|---|
| `account_class` | `classe_conta` | Yes | `5` or `6` |

**Account classes:**

- `5` -- Budgetary planning and control - approved (Controle da aprovacao)
- `6` -- Budgetary execution (Execucao do orcamento)

```{r}
# English -- budgetary execution at year-end
msc_exec <- get_msc_budget(
  entity_id     = 26,
  year          = 2023,
  month         = 12,
  matrix_type   = "MSCC",
  account_class = 6,
  value_type    = "period_change"
)

# Portuguese (equivalent)
msc_exec <- get_msc_orcamentaria(
  id_ente        = 26,
  an_referencia  = 2023,
  me_referencia  = 12,
  co_tipo_matriz = "MSCC",
  classe_conta   = 6,
  id_tv          = "period_change"
)
```

**Key response columns:** Same as MSC equity, plus `natureza_despesa`,
`natureza_receita`, `funcao`, `subfuncao`.

---

### 9. MSC control accounts -- `get_msc_control()` / `get_msc_controle()`

**Endpoint:** `/msc_controle`

Same parameters as MSC equity, except:

| English param | Portuguese param | Required | Valid values |
|---|---|---|---|
| `account_class` | `classe_conta` | Yes | `7` or `8` |

**Account classes:**

- `7` -- Off-balance-sheet controls - debit (Controles devedores)
- `8` -- Off-balance-sheet controls - credit (Controles credores)

```{r}
# English -- control accounts at year-end
msc_ctrl <- get_msc_control(
  entity_id     = 26,
  year          = 2023,
  month         = 12,
  matrix_type   = "MSCC",
  account_class = 8,
  value_type    = "ending_balance"
)

# Portuguese (equivalent)
msc_ctrl <- get_msc_controle(
  id_ente        = 26,
  an_referencia  = 2023,
  me_referencia  = 12,
  co_tipo_matriz = "MSCC",
  classe_conta   = 8,
  id_tv          = "ending_balance"
)
```

**Key response columns:** Same as MSC equity, plus `natureza_despesa`,
`funcao`, `subfuncao`.

---

## Quick reference: common IBGE codes

Some frequently used IBGE codes for `entity_id` / `id_ente`:

| Entity | IBGE code | Sphere |
|---|---|---|
| Federal Government | `1` | `"U"` |
| Sao Paulo (state) | `35` | `"E"` |
| Minas Gerais (state) | `31` | `"E"` |
| Rio de Janeiro (state) | `33` | `"E"` |
| Pernambuco (state) | `26` | `"E"` |
| Sao Paulo (city) | `3550308` | `"M"` |
| Rio de Janeiro (city) | `3304557` | `"M"` |
| Recife (city) | `2611606` | `"M"` |
| Belo Horizonte (city) | `3106200` | `"M"` |

Use `get_entities()` to look up any entity's IBGE code:

```{r}
library(dplyr)

get_entities() |>
  filter(uf == "PE", capital == 1) |>
  select(cod_ibge, ente, populacao)
```

---

## Tips

- Use `get_annexes()` to discover valid `appendix` values before calling
  report functions.
- Use `get_delivery_status()` to verify that a report has been submitted
  before querying it.
- All functions accept `use_cache = FALSE` to bypass the in-memory cache.
- Call `siconfir_clear_cache()` to reset cached data during a session.
